---
output:
  html_document:
    keep_md: yes
---
# Reproducible Research: Assignment 1

## Loading and preprocessing the data
To do the offline process, this implementation has downloaded the file to be accessed locally.
```{r}
ActivityData <- read.csv("activity.csv")
str(ActivityData)
```

## Problem 1: What is the mean total number of steps taken per day?
The following histogram displays the total number of steps taken per day, ignoring missing values.
```{r}
ActivityData.date <- aggregate(ActivityData[1],by=ActivityData[2],FUN=sum, na.rm=TRUE)
hist(ActivityData.date$steps,
     breaks=20,
     col = "green",
     main = "Histogram of Total Number of Steps Taken per Day",
     xlab = "Steps per Day")

mean(ActivityData.date$steps)   # mean
median(ActivityData.date$steps) # median

```

**Mean:** Ignoring missing values, the mean number of steps taken per day is `r format(mean(ActivityData.date$steps),big.mark=',')`.

**Median:** Ignoring missing values, the median number of steps taken per day is `r format(median(ActivityData.date$steps),big.mark=',')`.

## Problem 2:  What is the average daily activity pattern?
Following is a time-series plot of the average number of steps taken during each 5-minute interval during the day.
```{r}
# The following data set contains the sample data aggregated by interval.
ActivityData.interval <- aggregate(ActivityData[1],by=ActivityData[3],FUN=mean,na.rm=TRUE)
plot(x=ActivityData.interval$interval,
     y=ActivityData.interval$steps,
     type="l",
     main="Average Steps Per 5-Minute Interval",
     xlab="Interval",
     ylab="Number of Steps")
```

**Maximum Steps and Interval it occurred**

```{r}
Maximum_Interval <- ActivityData.interval[ActivityData.interval$steps==max(ActivityData.interval$steps),]

Maximum_Interval[1] # maximum interval
round(Maximum_Interval[2],1) # maximum steps per interval
```

The maximum average steps per period occur at interval `r Maximum_Interval[1]`, with an average of `r round(Maximum_Interval[2],1)` steps per 5 minutes (calculated inline).

## Imputing missing values
**Step 1 - Number of Missing Values**
```{r}
# total records
nrow(ActivityData)
# missing records
sum(is.na(ActivityData$steps))
```
The total number of records in the dataset is `r format(nrow(ActivityData),big.mark=',')`.  Of these, `r format(sum(is.na(ActivityData$steps)),big.mark=',')` contain missing values for number of steps taken.

**Step 2 and 3 - Replacing Missing Values and Creating a New Dataset**

Missing values for steps per interval are replaced by the mean number of steps for that interval, calculated on the non-missing rows.
```{r}
library(plyr)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
ActivityData.impute <- ddply(ActivityData, ~interval, transform, steps = impute.mean(steps))
```

**Step 4 - Histogram, Mean and Median**

The following is a histogram of the total number of steps per day using imputed values for missing values.
```{r}
# The following data set contains the sample data aggregated by date.
ActivityData.impute.date <- aggregate(ActivityData.impute[1],by=ActivityData.impute[2],FUN=sum,na.rm=TRUE)
hist(ActivityData.impute.date$steps,
     breaks=20,
     col = "green",
     main = "Histogram of Total Number of Steps Taken per Day (Imputed Data)",
     xlab = "Steps per Day")

mean(ActivityData.impute.date$steps)   # mean
median(ActivityData.impute.date$steps) # median

```

**Mean:** Ignoring missing values, the mean number of steps taken per day, rounded to the nearest step, is `r format(mean(ActivityData.impute.date$steps),big.mark=',')`.

**Median:** Ignoring missing values, the median number of steps taken per day, rounded to the nearest step, is `r format(median(ActivityData.impute.date$steps),big.mark=',')`.

This means of value imputation has increased both the mean and the median values from those calculated in the first part of the assignment. 

## Problem 3:  Are there differences in activity patterns between weekdays and weekends?
**Step 1 - New Factor Variable**

A new factor variable, Day, is created, indicating whether a record pertains to a weekday or to a weekend.
``` {r}
ActivityData.impute$dateP <- as.POSIXlt(ActivityData.impute$date,format="%Y-%m-%d")
ActivityData.impute$day <- "Weekday"
ActivityData.impute$day [weekdays(ActivityData.impute$dateP) %in% c("Saturday","Sunday")] <- "Weekend"
```

**Step 2 - Weekday/Weekend Comparison**

The following time series plots display the weekday and weekend data separately.  

``` {r}
ActivityData.impute.interval <- aggregate(ActivityData.impute[1],
                                   by=ActivityData.impute[c(3,5)],
                                   FUN=mean,
                                   na.rm=TRUE)
library(ggplot2)
plot <- ggplot(data = ActivityData.impute.interval,
               aes(x=interval,y=steps))
plot + geom_line() + facet_wrap(~day,nrow=2)

```

**Summary**
From the chart, it looks like people were active during weekend.
